---
  title: "AnforFree algorithm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{AnchorVolume}
%\VignetteEngine{knitr::rmarkdown}
%\VignetteEncoding{UTF-8}
---
  
  
  
  Generally, structured matrix factorization problem is formulated as a decomposition of original matrix $X$ in a product of unknown non-negative matrix $C$ and matrix $D$ of lower rank: 
  
  $$
  X = C \cdot D
$$
  In practice, if $X$ has significantly more rows than columns or if observations are noisy and represent statistical sampling than reformulation of the problem in "covariance domain" simplifies inference:
  
  $$
  Y = C \cdot E \cdot C^{T},
$$
  where $Y = X \cdot X^{T}, E = D \cdot D^{T}.$
  
  The goal of _AnchorFree_ approach is to tri-factorize matrix $Y$ in a product of matrices $C$ and $E$. Under a relatively mild assumption on spread of column vectors of matrix $C$, matrix $E$ has minimum volume across all possible factorizations. _AnchorFree_ seeks to find a factorization pair of matrices $(C,E)$ such that column vectors of $C$ represent unit simplex and matrix $E$ has minimum determinant (as a proxy of volume) using alternating linear programming. 

Application of the method is examplified below for a simulated dataset. For that, we will use dataset with matrices `X`, its noisy version `X.noise` and decomposition matrices `C` and `D` such that `X = C*D` (to simulate a dataset with parameters of interest one can use function `sim_factors`). First, we upload the library and the dataset:
  
  ```{r}
library(vrnmf)

data(simnmf)

names(simnmf)
```

Summary below shows that the original matrix `X` of the size $192 \cdot 3000$ is a product of two matrices of lower rank 10:
  
  ```{r}
str(simnmf$X)

str(simnmf$C)

str(simnmf$D)
```

As a first step, input matrix `X` is preprocessed using function `vol_preprocess()` to estimate and re-scale co-occurence matrix `Y = X*t(X)` and its singular value decomposition:
  
  ```{r}
vol <- vol_preprocess(t(simnmf$X), pfactor = 1e+5)
```

A parameter `pfactor` re-scales co-occurence matrix, but does not affect outcome. Currently, due to technical issues in implementation, too small `pfactor` can collapse volume to zero and we recommend to increase the factor to avoid numerical issues.

Next, `AnchorFree` function decomposes co-occurence matrix into matrices `C` and `E` of rank `n.comp` by minimizing volume of `E`:
  
  ```{r}
vol.anchor <- AnchorFree(vol, n.comp = 10)

str(vol.anchor$C)

str(vol.anchor$E)
```

Comparison of original matrix `simnmf$C` and inferred matrix `vol.anchor$C` shows that they have highly similar pairs of column vectors:
  
  ```{r}
C <- vol.anchor$C
apply(cor(simnmf$C, C), 1, max)
```

Having inferred `C`, the second matrix `D` can be obtained throught linear regression with constraints or using the function `infer_intensities()`:
  
  ```{r}
D <- infer_intensities(C, simnmf$X)
```

